Triangulating iconicity: From structure mapping to guessability

### 2020-06-26 TI_02_congruence.Rmd
tinkering with visualizations and gghalves

```{r preliminaries, results="hide"}
# Packages
list.of.packages <- c("tidyverse","readxl","writexl","ggthemes","viridis","lme4","VGAM")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only=T)

@@

@@

* TO DISCUSS: We need to think about the evidential value of each of the congruency measures. Most apply to many words, which is good. If some of them only capture a few (as `C_durative` and `C_weight_tone` seem to do), they are not so interesting because they don't really capture recurring iconic mappings in the data. Or perhaps this may indicate a need to recode — e.g. perhaps `F_finallength` should not just map onto aspectual semantics (like `M_durative` but also to shape, sth like `M_long`: a very concrete kind of mapping.)

@@

# using ifelse() statements that essentially say, if these conditions are true, use 1, otherwise 0
d <- d %>%
  mutate(C_modality = ifelse(M_sound == 1,1,0),
         C_iterative = ifelse(F_redup == 1 & M_distribution == 1,1,0),
         C_irregular = ifelse(F_redupmod == 1 & M_irregular == 1,1,0),
         C_closure = ifelse(F_closedsyllable == 1 & M_abrupt == 1,1,0),
         C_punctual = ifelse(F_monosyllabic == 1 & M_punctual == 1,1,0),
         C_durative = ifelse(F_finallength == 1 & M_durative == 1,1,0),
         C_weight_voice = ifelse((F_voice == 0 & M_weight == 0) | (F_voice == 2 & M_weight) == 2,1,0),
         C_weight_vowel = ifelse((F_vowelquality == 0 & M_weight == 0) | (F_vowelquality == 2 & M_weight == 2),1,0),
         C_weight_tone = ifelse((F_intonation == 0 & M_weight == 2) | (F_intonation == 2 & M_weight == 0),1,0)
         ) %>%
  mutate(C_cumulative = C_modality + C_iterative + C_irregular + C_closure + C_punctual + C_durative + C_weight_voice + C_weight_vowel + C_weight_tone) %>%
  mutate(C_any = ifelse(C_cumulative == 0,0,1),
         C_atleast2 = ifelse(C_cumulative > 1,1,0),
         C_ternary = ifelse(C_cumulative == 0,0,
                            ifelse(C_cumulative == 1,1,2)))

d %>%
  group_by(C_cumulative) %>%
    summarise(n=n())

@@

# all data points by language
icoplot()

icoplot("C_modality")
icoplot("C_iterative")
icoplot("C_irregular")
icoplot("C_punctual")
icoplot("C_durative")
icoplot("C_weight_voice")
icoplot("C_weight_vowel")
icoplot("C_weight_tone")

@@

icoplot("C_any")
icoplot("C_atleast2")
icoplot("C_ternary")

@@

## Exploring congruency blind spots
What are we not capturing? Our coding and congruency measures are designed to be reproducible and human-codable. They are not exhaustive, and there may well be aspects that we are missing. Here we explore the congruency data to find highly guessable ideophones that our coding did not (yet) identify as having iconic mappings.

@@

d.blindspots <- d %>%
  filter(C_cumulative == 0, score >0.5) %>%
  dplyr::select(filename,ideophone,meaning,meaning_NL,language,score) %>%
  arrange(-score)
View(d.blindspots)

@@

## Stats

@@

cor.test(d$logodds,d$C_cumulative)
cor.test(d$score,d$C_cumulative)

>>>>

```{r preliminaries, results="hide"}
# Packages
list.of.packages <- c("tidyverse","readxl","writexl","ggthemes","gghalves","ggbeeswarm","viridis","lme4","VGAM")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only=T)

@@

# add Z score to account for differences in base success rate of Collabra and Language Studies
d <- d %>%
  group_by(study) %>%
  mutate(score_z = scale(score,center=T,scale=T))

@@

* TO DISCUSS: We need to think about the evidential value of each of the congruency measures. Most apply to many words, which is good. If some of them only capture a few (as `C_durative` and `C_weight_tone` seem to do), they are not so interesting because they don't really capture recurring iconic mappings in the data. 
* Discussed: This made us realize a need to name `M_durative` to `M_long`, covering extent not just in time but also in space (a very concrete kind of mapping for `F_finallength`.

@@

# using ifelse() statements that essentially say, if these conditions are true, use 1, otherwise 0
d <- d %>%
  mutate(C_modality = ifelse(M_sound == 1,1,0),
         C_iterative = ifelse(F_redup == 1 & M_distribution == 1,1,0),
         C_irregular = ifelse(F_redupmod == 1 & M_irregular == 1,1,0),
         C_closure = ifelse(F_closedsyllable == 1 & M_abrupt == 1,1,0),
         C_punctual = ifelse(F_monosyllabic == 1 & M_punctual == 1,1,0),
         C_long = ifelse(F_finallength == 1 & M_long == 1,1,0),
         C_weight_voice = ifelse((F_voice == 0 & M_weight == 0) | (F_voice == 2 & M_weight) == 2,1,0),
         C_weight_vowel = ifelse((F_vowelquality == 0 & M_weight == 0) | (F_vowelquality == 2 & M_weight == 2),1,0),
         C_weight_tone = ifelse((F_intonation == 0 & M_weight == 2) | (F_intonation == 2 & M_weight == 0),1,0)
         )

# are some mappings mutually exclusive? then it might be slightly more elegant to combine them
d %>%
  mutate(C_event = C_iterative + C_punctual) %>%
  dplyr::select(ideophone,meaning_NL,language,C_iterative,C_punctual,C_event) %>%
  filter(C_event > 1)

d %>%
  mutate(C_exclusive = C_long + C_closure) %>%
  dplyr::select(ideophone,meaning_NL,language,C_iterative,C_closure,C_exclusive) %>%
  filter(C_exclusive > 1)

# add some cumulative iconicity scores
d <- d %>%
  mutate(C_cumulative = C_modality + C_iterative + C_irregular + C_closure + C_punctual + C_long + C_weight_voice + C_weight_vowel + C_weight_tone) %>%
  mutate(C_any1 = ifelse(C_cumulative == 0,0,1),
         C_atleast2 = ifelse(C_cumulative > 1,1,0),
         C_ternary = ifelse(C_cumulative == 0,0,
                            ifelse(C_cumulative == 1,1,
                                   ifelse(C_cumulative == 2,2,3))))

# cumulative scores can be further refined by considering only features that work for this sample — e.g., since C_irregular only adds noise, it may be excluded 
d <- d %>%
  mutate(C_simple = C_modality + C_iterative + C_closure + C_punctual + C_weight_tone + C_weight_vowel)

# write this dataset for use in the next step
write.csv(d,file="data/ideophones_coded_guessed.csv")

d %>%
  group_by(C_cumulative) %>%
  summarise(n=n(),score=mean.na(score))

d %>%
  group_by(C_ternary) %>%
  summarise(n=n(),score=mean.na(score))

d %>%
  group_by(C_simple) %>%
  summarise(n=n(),score=mean.na(score))

@@

# all data points by language
icoplot() # icoplot() is loaded from TI_functions.R

icoplot("C_modality")
icoplot("C_iterative")
icoplot("C_irregular")
icoplot("C_punctual")
icoplot("C_long")
icoplot("C_weight_voice")
icoplot("C_weight_vowel")
icoplot("C_weight_tone")

@@

icoplot("C_any1")
icoplot("C_atleast2")
icoplot("C_ternary")

@@

### Tinkering with plots
Just trying different ways of visualizing the data to find the most helpful and transparent one.

```{r visuals}

icoplot("C_simple")

ggplot(data=dp, aes(x="all",y=score_z)) +
  theme_tufte() +
  geom_half_boxplot() +
  geom_half_point(side="l") +
  geom_half_violin(side="r") +
  NULL

ggplot(data=dp, aes(x=C_simple,y=logodds)) +
  theme_tufte() +
  geom_half_boxplot() +
  geom_half_point() +
  NULL
ggplot(data=dp, aes(x=C_simple,y=score_z)) +
  theme_tufte() +
  geom_half_boxplot() +
  geom_half_point() +
  NULL

ggplot(data=dp, aes(x=C_simple,y=logodds,color=C_simple)) +
  theme_tufte() +
  geom_half_boxplot() +
  geom_half_dotplot() +
  scale_fill_viridis(option="plasma",discrete=T,begin=0.3,end=0.9) +
  scale_colour_viridis(option="plasma",discrete=T,begin=0.3,end=0.9) +
  NULL

ggplot(data=dp, aes(x=C_ternary,y=logodds,color=C_ternary,fill=C_ternary)) +
  theme_tufte() +
  geom_half_violin() +
  geom_half_point_panel() +
  scale_fill_viridis(option="plasma",discrete=T,begin=0.3,end=0.9) +
  scale_colour_viridis(option="plasma",discrete=T,begin=0.3,end=0.9) +
  facet_grid(~ study)

ggplot(data=dp, aes(x=C_ternary,y=logodds,color=C_ternary,fill=C_ternary)) +
  theme_tufte() +
  geom_half_violin() +
  geom_half_dotplot(dotsize=0.8) +
  scale_fill_viridis(option="plasma",discrete=T,begin=0.3,end=0.9) +
  scale_colour_viridis(option="plasma",discrete=T,begin=0.3,end=0.9) +
  facet_grid(~ study)

ggplot(data=dp, aes(x=C_atleast2,y=logodds,color=C_atleast2,fill=C_atleast2)) +
  theme_tufte() +
  geom_half_violin() +
  geom_half_dotplot(dotsize=0.4,stackratio=0.8,method="histodot") +
  scale_fill_viridis(option="plasma",discrete=T,begin=0.3,end=0.9) +
  scale_colour_viridis(option="plasma",discrete=T,begin=0.3,end=0.9) +
  facet_grid(~ study)

ggplot(data=dp, aes(x=category,y=logodds,color=category)) +
  theme_tufte() +
  geom_half_violin() +
  geom_half_point_panel() +
  NULL

ggplot(data=dp, aes(x=language,y=logodds,color=language)) +
  theme_tufte() +
  geom_half_violin() +
  geom_half_point_panel() +
  NULL

ggplot(data=d, aes(x=language,y=C_ternary,color=language)) +
  theme_tufte() +
  geom_half_violin() +
  geom_half_point_panel() +
  NULL

## Exploring congruency blind spots
What are we not capturing? Our coding and congruency measures are designed to be reproducible and human-codable. They are not exhaustive, and there may well be aspects that we are missing. Some ideophones may also be guessed relatively high for reasons other than form-meaning analogies (e.g., they had a very unattractive foil or presented unforeseen confounds). Here we explore the congruency data to find highly guessed ideophones that our coding did not (yet) identify as having iconic mappings. We also look at the other end: ideophones coded as having an iconic mapping but not doing well in the binary forced choice task.

If we sort by score most of the blindspots are from the Collabra study, which had a less forbidding design and therefore a higher baseline score.

@@

d %>%
  filter(C_cumulative == 0, logodds >0) %>%
  dplyr::select(filename,ideophone,meaning,meaning_NL,language,logodds,score) %>%
  arrange(-score) %>% ungroup %>%
  slice(1:10)
```

```{r congruency_underperformers}
# the reverse is also interesting: which ideophones were coded as having iconic mappings yet were not guessed greatly?

d %>%
  filter(C_cumulative > 0, logodds < -0.25) %>%
  dplyr::select(filename,ideophone,meaning,meaning_NL,C_cumulative,language,logodds,score) %>%
  arrange(score) %>% ungroup() %>%
  slice(1:10)

@@

## Stats
As the visualizations already make clear there is a non-trivial positive relation between the presence of iconic mappings (in a cumulative sense) and the guessability score determined in experimental work.

@@

cor.test(d$logodds,d$C_cumulative)
cor.test(d$score_z,d$C_cumulative)

### 2020-06-26 TI_03_ratings.Rmd
tinkering with visualizations and gghalves

Code notebook for a study of the relation between linguistically informed iconicity coding and experimentally collected guessability scores.

@@

```{r preliminaries, results="hide"}
# Packages
list.of.packages <- c("tidyverse","readxl","writexl","ggthemes","viridis")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only=T)

@@

>>>>

Code notebook for a study of the relation between iconicity ratings and experimentally collected guessability scores.

@@

```{r preliminaries, results="hide"}
# Packages
list.of.packages <- c("tidyverse","readxl","writexl","ggthemes","viridis","lme4","VGAM")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only=T)

@@

## Data
Here we load the combined coding and guessability data and then add the rating data.

```{r data}
# get coding data
d = read_excel("data\\ideophones_coded.xlsx") %>% arrange(filename)

coding_categories <- d %>% dplyr::select(matches('_'),-matches('notes|meaning')) %>% names()

# add guessability scores from the Collabra and Language studies.

d.scores = read_excel("data\\ideophones_guessability.xlsx") %>%
  dplyr::select(-category)
d <- left_join(d,d.scores,by=c("ideophone","language","study" = "paper"))

# add logodds (for when we run stats: it's more sensible to predict against logodds than raw proportion correct)
d <- d %>%
  group_by(study) %>%
  mutate(logodds = probitlink(score))

# add mean ratings
d.ratings = read_excel("data\\ideophones_rated_means.xlsx") %>%
  dplyr::select(-c(item,category,list))
d <- left_join(d,d.ratings,by=c("filename","language","study"))

# convert ratings to z-scores for comparison with guessability (logodd) scores
d$rating_z <- scale(d$rating,center=T,scale=T)

## Statistics
```{r stats}
# before calculating the correlation coefficient, we visualise the correlation between the guessing and rating data
ggplot(d,aes(x=rating_z,y=logodds)) + geom_point(colour="dark grey",position="jitter") + geom_smooth(method=lm,colour="black")
ggsave("figures\\TI_03_corr.png",height=5,width=7.5)

ggplot(d,aes(x=rating_z,y=logodds)) + geom_point(aes(color=category,shape=language),position="jitter")
ggsave("figures\\TI_03_corr_bylanguageandcategory.png",height=5,width=7.5)

# Pearson correlation
cor.test(d$logodds,d$rating_z,method="pearson")

Pearson's correlation coefficient indicates that there is a strong positive correlation between the guessing and rating scores, r(237) = .57, p < .001.
